On selection of a coupling method to cope with change of priors in one-vs-one classification

### 2025-05-26 coupling.tex
Various fixes

Suppose first that real numbers $a_1, \ldots,a_M$ are real numbers and $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_M$ are probabibilistic multi-class classification methods satisfying Assumption \ref{ass:2}. We define their linear combination $\bigoplus_i a_m \boldsymbol{v}_m$ as the classification method that for any $x$ yields the probability distribution on $K$ classes that satisfies
\begin{align*}
 	\bigl(a_1 \boldsymbol{v}_1 \oplus \ldots \oplus a_M \boldsymbol{v}_M\bigr)(x) \propto \boldsymbol{v}_1^{a_1}(x) \odot \cdots \odot \boldsymbol{v}_M^{a_M}(x),
\end{align*}
where the symbol $\odot$ denotes componentwise multiplication, and exponentials are computed component-wise. With this definition we can state the following result.

@@

It follows from the definition of linear combination of coupling methods that the odds of $C_i$ to $C_j$ for $\boldsymbol{e}$ are
\begin{align}
\prod_m o_m^{a_m}.
\end{align}
It follows that the odds of $\boldsymbol{e}^\pi$ are 
\begin{align}
\frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}.
\end{align}

Let us now use Bayes covariance of $\boldsymbol{v}_i$. 
We have
\begin{align}
	\biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)(\boldsymbol{R^\pi}) & \propto 
	\boldsymbol{v}_1^{a_1}(\boldsymbol{R}^\pi) \odot \cdots \odot \boldsymbol{v}_M^{a_M}(\boldsymbol{R}^\pi) =  \bigodot_m  (\boldsymbol{v}_m^\pi(\boldsymbol{R}))^{a_m} \\
%	&= \frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}
\end{align}

It follows that the odds of $C_i$ to $C_j$ for $\bigoplus_m a_m \boldsymbol{v}_m$ are 
\begin{align}
\biggl(\frac{P'_i}{P_i} \frac{P_j}{P'_j} \biggr)^{\sum a_m} \prod_m o_m^{a_m}.
\end{align}
Clearly, under the assumption $\sum_m a_m = 1$, the classifier $\bigoplus_m a_m \boldsymbol{v}_m$ satisfies $BC_{ij}$. Since this holds for any $i\not=j$,  it follows that $\boldsymbol{e}$ is Bayes covariant.
\end{proof}

>>>>

Suppose first that $a_1, \ldots,a_M$ are real numbers and $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_M$ are probabilistic multi-class classification methods satisfying Assumption \ref{ass:2}. We define their linear combination $\bigoplus_i a_m \boldsymbol{v}_m$ as the classification method that for any $x$ yields the probability distribution on $K$ classes that satisfies
\begin{align*}
 	\bigl(a_1 \boldsymbol{v}_1 \oplus \ldots \oplus a_M \boldsymbol{v}_M\bigr)(x) \propto \boldsymbol{v}_1^{a_1}(x) \odot \cdots \odot \boldsymbol{v}_M^{a_M}(x),
\end{align*}
where the symbol $\odot$ denotes componentwise multiplication, and exponentials are computed component-wise. With this definition we can state the following result.

@@

From the definition of linear combination of coupling methods that the odds of $C_i$ to $C_j$ for $\boldsymbol{e}$ are
\begin{align}
\prod_m o_m^{a_m}.
\end{align}

It follows from \eqref{def:prior:effect} that the odds of $\boldsymbol{e}^\pi$ are 
\begin{align}
\frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}.
\end{align}

Let us now use Bayes covariance of $\boldsymbol{v}_i$. 
We have
\begin{align}
	\biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)(\boldsymbol{R^\pi}) & \propto 
	\boldsymbol{v}_1^{a_1}(\boldsymbol{R}^\pi) \odot \cdots \odot \boldsymbol{v}_M^{a_M}(\boldsymbol{R}^\pi) =  \bigodot_m  (\boldsymbol{v}_m^\pi(\boldsymbol{R}))^{a_m}. \\
%	&= \frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}
\end{align}
Therefore the odds of $C_i$ to $C_j$ for $\bigoplus_m a_m \boldsymbol{v}_m$ are 
\begin{align}
\biggl(\frac{P'_i}{P_i} \frac{P_j}{P'_j} \biggr)^{\sum a_m} \prod_m o_m^{a_m}.
\end{align}

Clearly, under the assumption $\sum_m a_m = 1$, the classifier $\bigoplus_m a_m \boldsymbol{v}_m$ satisfies $BC_{ij}$. Since this holds for any $i\not=j$,  it follows that $\boldsymbol{e}$ is Bayes covariant.
\end{proof}

### 2025-05-26 experiment.tex
Fix Figure 2 (from 1st experiment)
Various fixes

@@

\input{tab-multi.tex}

From Table \ref{tab:multi} we can clearly see that Wu-Lin-Weng's method has the best performance. On some datasets it trounces normal coupling method by a wide margin. The edge over radial method is smaller, with the exception of usps dataset where its accuracy is higher by remarkable 30\%.

\begin{figure}
	\includegraphics{graphs/exp1-plot1.pdf}
	\caption{0-1 score plotted against the logarithm of $tol$ hyper-parameter, evaluated for different datasets.}
	\label{fig:exp1-plot1}
\end{figure}

@@

In this experiment we investigated whether we can (always) find  a Bayes covariant coupling method that could match the performance of Wu-Lin-Weng's method, if we allow it to have (a few) trainable parameters.  Our experiment proceeds in five steps.

First, we restrict our investigation to the case of three class datasets. This is a natural starting point, since it is the smallest number of classes where the question is meaningful to ask.

As a second step, we consider whether the three arboreal methods $ \boldsymbol{s}_1,~ \boldsymbol{s}_2,~ \boldsymbol{s}_3$ (which do not have any parameters) could bridge the gap. The statistics are shown in Table \ref{tab:step2}.

@@

We can see that approximately half of the cases none of the four parameterless Bayes covariant methods was able to match the performance of Wu-Lin-Weng's method. Therefore we should extend search to Bayes covariant methods with trainable parameters.

In the third step we define two parameter family of Bayes covariant coupling methods
\begin{align}
a_1 \boldsymbol{s}_1 \oplus a_2 \boldsymbol{s}_2 \oplus a_3 \boldsymbol{s}_3,\quad\textrm{where } a_1 + a_2 +a_3 = 1 \label{eq:family}
\end{align}
Note that the parameter space of this family is not compact. In order to facilitate computations we opted to conduct a grid search on points in the simplex where the coefficients $a_i$ are nonnegative $a_i \geq 0$.  To further reduce computational burden we selected 1000 three-class subproblems from those where Wu-Lin-Weng's method outperformed normal coupling. The results are shown in Figure \ref{fig:par-bc}.

>>>>

\section{Methodology.}

@@

% \input{tab-multi.tex}

From Figure \ref{fig:exp1-plot1} we can clearly see that Wu-Lin-Weng's method has the best performance. On some datasets it trounces normal coupling method by a wide margin. The edge over radial method is smaller, with the exception of usps dataset where its accuracy is higher by remarkable 30\%.

\begin{figure}
	\includegraphics{graphs/exp1-plot2.pdf}
	\caption{0-1 score plotted against the logarithm of $tol$ hyper-parameter, evaluated for different datasets.}
	\label{fig:exp1-plot1}
\end{figure}

@@

In this experiment we investigated whether we can (always) find  a Bayes covariant coupling method that could match the performance of Wu-Lin-Weng's method. We investigate  the case of three class datasets. This is a natural starting point, since it is the smallest number of classes where the question is meaningful to ask if we allow it to have (a few) trainable parameters.  Our experiment has three parts.

In the first part, we consider whether four parameterless Bayes covariant method could match the performance of Wu-Lin-Weng method. In our comparison we include the normal method, as well as the three arboreal methods $ \boldsymbol{s}_1,~ \boldsymbol{s}_2,~ \boldsymbol{s}_3$. The statistics are shown in Table \ref{tab:step2}.

@@

We can see that in slightly less than half of the cases none of the four parameterless Bayes covariant methods was able to match the performance of Wu-Lin-Weng's method. Therefore next we will extend search to Bayes covariant methods with trainable parameters.

Let us define two parameter family of Bayes covariant coupling methods
\begin{align}
a_1 \boldsymbol{s}_1 \oplus a_2 \boldsymbol{s}_2 \oplus a_3 \boldsymbol{s}_3,\quad\textrm{where } a_1 + a_2 +a_3 = 1 \label{eq:family}
\end{align}
Note that the parameter space of this family is not compact.

In the second part of our experiment we  conduct a grid search on points in the simplex where the coefficients $a_i$ are nonnegative $a_i \geq 0$.  To further reduce computational burden we selected 1000 three-class subproblems from those where Wu-Lin-Weng's method outperformed normal coupling. The summary statistics are shown in Figure \ref{fig:par-bc}.

### 2025-05-26 paper.tex
Various fixes

\title{On change of priors in one-vs-one classification methods}

\author{\name Ondrej Šuch \email ondrejs@savbb.sk \\
       \addr Matematický ústav SAV\\
       Ďumbierska 1\\
       Banská Bystrica, 974 01, Slovakia
       \AND
       \name Peter Novotný \email peter.novotny@uniza.sk \\
       \addr Fakulta informatiky a riadenia\\
       Žilinská Univerzita v Žiline\\
       Žilina, 010 26, Slovakia
       \AND
       \name Ali Haidar \email haidar@savbb.sk \\
       \addr Matematický ústav SAV\\
       Ďumbierska 1\\
       Banská Bystrica, 974 01, Slovakia
       }

@@

Multi-class classification is considered to be a more challenging problem than binary classification.  A natural approach to multi-class classification is to reduce it to a series of binary classification problems and deduce the multi-class decision by aggregating decisions for the binary problems.

One popular method belonging to  this paradigm uses probabilistic modelling for one-vs-one series of binary problems. Notably, it is used as the basis for multi-class classification models using support vector machines (SVM) in LIBSVM library. However one can use the approach more generally,  because it can be applied to any probabilistic binary classification method.

A key step in probabilistic one-vs-one modelling is the aggregation of results of individual binary classifiers - so-called \emph{coupling method}. The most commonly used coupling method is that proposed by Wu-Lin-Weng. As noted in [\cite{dohau}], the method is one of several non-canonical decisions used in multi-class SVM modelling. The goal of this paper is to compare its behavior with canonical methods.

@@

>>>>

\title{On selection of a coupling method to cope with change of priors in one-vs-one classification }

\author{\name Ondrej Šuch \email ondrejs@savbb.sk \\
       \addr Matematický ústav SAV\\
       Ďumbierska 1\\
       Banská Bystrica, 974 01, Slovakia
       \AND
       \name Peter Novotný \email peter.novotny@uniza.sk \\
       \addr Fakulta informatiky a riadenia\\
       Žilinská Univerzita v Žiline\\
       Žilina, 010 26, Slovakia
       \AND
       \name Ali Haidar \email haidar@savbb.sk \\
       \begin{minipage}[t]{0.45\textwidth}
       \addr Matematický ústav SAV\\
       Ďumbierska 1\\
       Banská Bystrica, 974 01, Slovakia
       \end{minipage}\hfill
       \begin{minipage}[t]{0.45\textwidth}
       \addr Fakulta matematiky, fyziky a informatiky\\
       Univerzity Komenského \\
       Mlynská dolina F1 \\
       Bratislava, 842 48, Slovakia
       \end{minipage}
       }

@@

Distribution shift is an important problem facing machine-learning practitioners  [\cite{zhang2023dive, sugiyama2007covariate}].  One particular way a distribution shift may manifest in classification problems  is when the class priors change between the training set used to train a model, and the data distribution in deployment - the so called label shift [\cite{vsipka2022hitchhiker,lipton2018detecting}]. To make a concrete example, consider a driver assistance system recognizing traffic signs. Suppose a speed limit sign was maliciously altered to indicate 80mph limit instead of 30mph. To avoid being fooled by the alteration, the system could utilize priors. The distribution of traffic signs will change markedly when the car leaves highway and enters streets in a city. It is very unlikely to encounter 80mph limit in a city setting. By taking into account the city priors, a driver assistance is likely to make the proper classification. In general a good classification system will be able to improve its performance when supplied with a new set of class priors that correspond to the true distribution of classes.

Change of priors can be effectively dealt with in probabilistic classifiers. The basic idea is that Bayes theorem stipulates \emph{exactly} what should happen  to the prediction of the  Bayes classifier when priors change.  In practice we assume that a trained classifier closely approximates the Bayes classifier. Therefore we can make adjustment to our prediction by applying the same transformation of posterior as would happen with the Bayes classifier.

There is a subtle issue with this approach  when handling probabilistic classifiers arising in one-vs-one classification framework.   One-vs-one classification is a two-step framework to approach  multi-class classification. In the first step it reduces the multiclass problem to a series of binary classification problems. In the second step it deduces the multi-class decision by aggregating decisions for the binary problems. Since the procedure is two-step, there is a possibility to apply adjustment of priors either to the probabilistic distributions obtained after the first step, or after the second step. It is not clear what should be the preferred course of action and this problem motivates our paper.

The focus of efforts is in presenting and evaluating the methods to aggregate binary predictions. We formulate the underlying principles of one-vs-one probabilistic classification in Section 2. Section is 3 is devoted to presenting different classes that one

@@

\input discussion.tex
